---
title: "CUNY SPS MSDS Bridge Winter 2018 - R Programming Final Project"
author: "S. Deokinanan"
date: "January 13, 2018"
output:
  html_document: default
  word_document: default
---
***
#### Background

Body Mass Index (BMI) is an indicator of the amount of body fat of a person. It is a measure of ones body weight in relation to their height. It is a tool that can assess a person's risk to diseases such as obesity and being underweight. According to the National Heart, Lung, and Blood Institute, BMI values 

* less than 18.5 kg/m^2^ are considered underweight.
* from 18.5 kg/m^2^ to 24.9 kg/m^2^ are healthy.
* from 25.0 to less than 30.0 kg/m^2^ are considered to be overweight.
* greater than 30.0 kg/m^2^ are considered to be obese.

People with BMI above 25.0 kg/m^2^ have an increased risk of heart disease, type 2 diabetes, hypertension, cardiovascular disease, and many more health risk [1, 3].

Many people are concerned about their weight and health. Talking to their doctors, an exercise coach or even when researching on the internet for diet tips, BMI is very likely to be mentioned. Thus, people tend to check their themselves. However, when compared to measurements taken by trained professionals, discrepancies can occur when one takes these measurements themselves. One example of a source of inconsistency in measurements is the sensitivity and type of home-scales being used. Digital home-scales have been found to provide more accurate and consistent measurements than dial scales [4]. According to a systematic review by Gorber, S.C. *et al*, both men and women are overwhelmingly seen overestimating their height and underestimating their weight [2]. This underestimation of weight is more pronounced in women [1, 3]. As a result, incorrect measurements of the weight and height can lead to incorrect calculation of one's BMI.

**The research question**  
Therefore, this project aims answer:

1. How common is it that an individuals can be mislead by their BMI measurements because of incorrect measurements of either their weight or height or both?
2. Is there any trend between a self-reported BMI and the actual BMI, distinguished by the gender?

***
#### Retrieving the Dataset

The dataset used in this project is called Self-Reports of Height and Weight. The dataset *Davis.csv* was retrieved from http://vincentarelbundock.github.io/Rdatasets/ on 3rd January, 2019 and was then save in my Github as *BodyHW*.

**Note**  
Upon visual inspection of the dataset, ID # 12 had input for a weight of 166 kg with a height of 57 cm which is very peculiar, while their reported weight and height were 56 kg and 163 respectively, therefore I decided to swap these numbers, strongly convinced they were entered in the wrong columns.

```{r}
# Retrieving the dataset from my Github
theURL <- "https://raw.githubusercontent.com/greeneyefirefly/sps-msds-BodyHW/master/dataset.csv"
BodyHW <- read.csv (file = theURL, header = TRUE, sep = ",")
```

***
##### Summary of the Dataset

The dataset contains 200 unique observations with 5 variables:

* sex - A factor with levels: F, female; M, male.
* weight - Measured weight in kg.
* height - Measured height in cm.
* repwt - Reported weight in kg. This variable has some missing values.
* repht - Reported height in cm. This variable has some missing values

###### Missing values are excluded in calculations.

```{r}
head(BodyHW)
summary(BodyHW)
aggregate(BodyHW$weight, by=list(BodyHW$sex), FUN=mean, na.rm=TRUE)
aggregate(BodyHW$height, by=list(BodyHW$sex), FUN=mean, na.rm=TRUE)
aggregate(BodyHW$repwt, by=list(BodyHW$sex), FUN=mean, na.rm=TRUE)
aggregate(BodyHW$repht, by=list(BodyHW$sex), FUN=mean, na.rm=TRUE)

```

The dataset consist of 112 females and 88 males. The table below shows the actual and the reported measurements for females and males.

Measures           | Overall  | Females  | Males
-------------------|----------|----------|---------
Actual Weight      | 65.25 kg | 56.89 kg | 75.90 kg
Reported Weight    | 65.62 kg | 56.74 kg | 76.56 kg
*Weight difference*| -0.37 kg |  0.15 kg | -0.66 kg
Actual Height      | 170.6 cm | 164.7 cm | 178.0 cm
Reported Height    | 168.5 cm | 162.2 cm | 176.3 cm
*Height difference*|   2.1 cm |   2.5 cm |   1.7 cm

The table suggest that both reported weight and height for female are underestimated when compared to their actual measurements, which is consistent with the literature. Whereas, for males, only height is underestimated when compared to their actual height. 

From the scatter plots below, it is clearer that there are discrepancies among the reported measurements and the actual weight and height.
```{r}
library(ggplot2)

# A scatter plot of the self-reported weight and height of the participants
ggRHW <- ggplot(BodyHW, aes(x=repwt, y=repht, na.rm = TRUE))
ggRHW + geom_point(na.rm = TRUE) + labs(title="Self-Reported Measurements", x="Self-Reported Weight", y="Self-Reported Height") + geom_smooth(method='lm', na.rm = TRUE)

# A scatter plot of the actual weight and height of the participants
ggAHW <- ggplot(BodyHW, aes(x=weight, y=height))
ggAHW + geom_point() + labs(title="Actual Measurements", x="Weight", y="Height") + geom_smooth(method='lm')
```

***
#### Investigating the Body Mass Index
The formula for calculating the Body Mass Index is: $BMI = weight (kg) / height^{2} (m^{2})$

```{r}
# Converting centimeter into meter for the actual and the reported height
BodyHW[["mheight"]] <- BodyHW$height/100
BodyHW[["repmht"]] <- BodyHW$repht/100

# Calculating the BMI from the actual and the reported measurements
BodyHW[["BMI"]] <- BodyHW$weight/(BodyHW$mheight^2)
BodyHW[["repBMI"]] <- BodyHW$repwt/(BodyHW$repmht^2)
```
 
#### The variablity of the actual and self-reported BMI by gender.
Looking at the box plots for the both the actual and self-reported BMI, males presented to have a greater variability than females. Moreover, the median for the self-reported BMI by males has a noticeable shift when compared to actual BMI by males.

```{r}
ggrepBMI <- ggplot(BodyHW, aes(x=sex, y=repBMI, na.rm = TRUE))
ggrepBMI + geom_boxplot(na.rm = TRUE) + labs(title="BMI based on Self-Reported Measurements distributed by Sex", x="Sex", y="Self reported BMI") 

ggABMI <- ggplot(BodyHW, aes(x=sex, y=BMI))
ggABMI + geom_boxplot() + labs(title="BMI based on Actual Measurements distributed by Sex", x="Sex", y="Actual BMI") 
```

***
#### Identifying the health status for where the participants can be placed based on their BMIs calculated with the actual and self-reported measurements.

```{r}
# Categorizing the BMI calculated from the actual measurements.
BodyHW[["ActualCategory"]] <- ifelse(BodyHW$BMI < 18.50,"underweight", 
                              ifelse(BodyHW$BMI > 18.50 & BodyHW$BMI < 24.90, "healthy",
                              ifelse(BodyHW$BMI > 24.90 & BodyHW$BMI < 30.0, "overweight", "obese")))

# Categorizing the BMI calculated from the self-reported measurements.
BodyHW[["repCategory"]] <- ifelse(BodyHW$repBMI < 18.50,"underweight", 
                           ifelse(BodyHW$repBMI > 18.50 & BodyHW$repBMI < 24.90, "healthy",
                           ifelse(BodyHW$repBMI > 24.90 & BodyHW$repBMI < 30.0, "overweight", "obese")))
```

After determining the participant's health status, the histograms show that are many self-reported BMI that differences from the actual BMI significantly. 

```{r}
ggrepBMI1 <- ggplot(BodyHW, aes(x=repBMI, fill=repCategory, na.rm = TRUE))
ggrepBMI1 + geom_histogram(bins=40, na.rm = TRUE)  + labs(title="BMI based on Self-Reported Measurements", x="Self-reported BMI", y="Count")

ggABMI2 <- ggplot(BodyHW, aes(x=BMI, fill=ActualCategory))
ggABMI2 + geom_histogram(bins=40) + labs(title="BMI based on Actual Measurements", x="Actual BMI", y="Count")
```

As a result, the table below highlights that there could have been 8 reports of *healthy* BMI when in fact they are actually 3 *overweight* and 5 *underweight* individuals. Having the belief that one is health when in fact they are not, can lead to healthy risk. Other cases could have included 8 reports of *overweight* and 1 report of *underweight* when in fact they are *healthy*, and 2 reports of *obese* when they are *overweight.*

Moreover, the table reveals that males would have more likely report *overweight* when in fact they are *healthy*, and vice versa. Females, however, would have tend to report *healthy* when they are in fact *underweight.*

```{r}
rows<-which(BodyHW$ActualCategory!=BodyHW$repCategory)
table(BodyHW[rows, c("sex","repCategory","ActualCategory")])
```

***
#### Conclusion

In conclusion, 10% of the individuals could have been mislead by their BMI measurements because of incorrect measurements of either their weight or height or both. Three percent of males could have been under the impression that they are *overweight* when they are instead *healthy*, and 2% of the females could have been mislead to be *healthy* when they are instead *underweight*. However, this does not imply that, for example, an incorrect BMI of *obese* for *overweight*, is not as alarming. In such cases, individuals should still pay close attention to their consumption and exercise activities. This is also the same recommendation for individuals who had incorrect BMI of *underweight* but are *healthy*, individuals should remain determined to maintain a healthy status, well between 18.5 kg/m^2^ to 24.9 kg/m^2^, particularly those at the cut-offs between categories and especially those who are *obese*.

***
***
##### Works Cited

###### 1. Engstrom, J.L.; Paterson, S.A.; Doherty, A.; Trabulsi, M.; Speer, K.L. Accuracy of self-reported height and weight in women: An integrative review of the literature. J. *Midwifery Women's Health* 2003, 48, 338-345.  
###### 2. Gorber, S.C.; Tremblay, M.; Moher, D.; Gorber, B. A comparison of direct vs. self-report measures for assessing height, weight and body mass index: A systematic review. *Obes. Rev.* 2007, 8, 307-326.  
###### 3. National Heart, Lung, and Blood Insitute. *Aim for a Healthy Weight*. Retrieved January 4, 2019, https://www.nhlbi.nih.gov/health/educational/lose_wt/ 
###### 4. Yorkin, M.; Spaccarotella, K.; Martin-Biggers, J.; Quick, V.; Byrd-Bredbenner, C. Accuracy and consistency of weights provided by home bathroom scales. *BMC Public Health* 2013, 13, 1194.






